Fundamental Problems with the Evidence Base of Adolescent Depression Treatments.

### 2024-01-13 apples_and_oranges.qmd
updated the adjusted cohens d

@@

@@

```
### Analysis of all studies

@@

z_psy_con_vs_med_con <- 0
for(i in 1: length(list_model_1_no_wl)){
z_psy_con_vs_med_con[i] <- list_model_1_no_wl[[i]]$zval[condition=="psychotherapy_control"]

@@

#### Meta-analysis of baseline values

We next conducted a meta-analysis to test for differences in means at baseline.

@@

# Metanalysis of CDRS at baseline #

df_with_cdrs <-  df_appl_v_orange %>%          
  filter(instrument_name == "cdrs")

library(meta)
means_cdrs_baseline_control <- metamean(n = baseline_n_control,
                   mean = baseline_mean_control,
                   sd = baseline_sd_control,
                   studlab = new_study_id,
                   data = df_with_cdrs,
                   sm = "MRAW",
                   fixed = FALSE,
                   random = TRUE,
                   method.tau = "REML",
                   hakn = TRUE,
                   title = "CDRS Scores at Baseline")
summary(means_cdrs_baseline_control)

subgroup_cdrs_baseline_control <- update.meta(means_cdrs_baseline_control, 
            subgroup = psy_or_med, 
            tau.common = FALSE)

means_cdrs_baseline_active <- metamean(n = baseline_n_active,
                                        mean = baseline_mean_active,
                                        sd = baseline_sd_active,
                                        studlab = new_study_id,
                                        data = df_with_cdrs,
                                        sm = "MRAW",
                                        fixed = FALSE,
                                        random = TRUE,
                                        method.tau = "REML",
                                        hakn = TRUE,
                                        title = "CDRS Scores at Baseline")
#summary(means_cdrs_baseline_active)

subgroup_cdrs_baseline_active <- update.meta(means_cdrs_baseline_active, 
            subgroup = psy_or_med, 
            tau.common = FALSE)

# Metanalysis of HAMD at baseline #
df_with_hamd <-  df_appl_v_orange %>%          
  filter(instrument_name == "hamd")

library(meta)
means_hamd_baseline_control <- metamean(n = baseline_n_control,
                                        mean = baseline_mean_control,
                                        sd = baseline_sd_control,
                                        studlab = new_study_id,
                                        data = df_with_hamd,
                                        sm = "MRAW",
                                        fixed = FALSE,
                                        random = TRUE,
                                        method.tau = "REML",
                                        hakn = TRUE,
                                        title = "HAMD Scores at Baseline")
#summary(means_hamd_baseline_control)

subgroup_hamd_baseline_control <- update.meta(means_hamd_baseline_control, 
            subgroup = psy_or_med, 
            tau.common = FALSE)

means_hamd_baseline_active <- metamean(n = baseline_n_active,
                                       mean = baseline_mean_active,
                                       sd = baseline_sd_active,
                                       studlab = new_study_id,
                                       data = df_with_hamd,
                                       sm = "MRAW",
                                       fixed = FALSE,
                                       random = TRUE,
                                       method.tau = "REML",
                                       hakn = TRUE,
                                       title = "HAMD Scores at Baseline")
#summary(means_hamd_baseline_active)

subgroup_hamd_baseline_active <- update.meta(means_hamd_baseline_active, 
            subgroup = psy_or_med, 
            tau.common = FALSE)

@@

>>>>

###########7. Function for baseline SMD results
library(broom)
run_metareg_and_extract <- function(dataframe, model_1, condition) {
  # Run metareg
  inst_baseline_metareg <- rma(yi = baseline_mean,
                               sei = baseline_stand_error,
                               data = dataframe,
                               method = "ML",
                               mods = model_1,
                               test = "knha")

  # Extract coefficients from the list
  baseline_metareg_results <- tidy(inst_baseline_metareg)
  baseline_metareg_results$condition <- condition
  coefs_for_inst_baseline <- tidy(inst_baseline_metareg)$estimate

  # Calculate SMDs
  smds_for_inst_baseline <- c(coefs_for_inst_baseline[1], coefs_for_inst_baseline[-1] + coefs_for_inst_baseline[1])

  # Calculate upper and lower CIs
  upper_cis_for_inst_baseline <- smds_for_inst_baseline + 1.96 * tidy(inst_baseline_metareg)$std.error
  lower_cis_for_inst_baseline <- smds_for_inst_baseline - 1.96 * tidy(inst_baseline_metareg)$std.error

  # Create dataframe with results
  df_baseline_means_with_cis <- data.frame(
    baseline_smds = smds_for_inst_baseline,
    upper_cis_smds = upper_cis_for_inst_baseline,
    lower_cis_smds = lower_cis_for_inst_baseline,
    condition = condition
  )

  # Return the two dataframes
  return(list(baseline_metareg_results = baseline_metareg_results, df_baseline_means_with_cis = df_baseline_means_with_cis))
}

@@

We conducted a number of additional analyses to test for the robustness of our results.

First, we tested whether the simulated values for the standard error had a substantial influence on the estimation of the differences between the medication vs psychotherapy control condition. To inspect whether this is the case, we plotted the z-value of the difference between the two coefficients against the number of simulations. We make inference on the stability of the difference, by counting the proportion of times that the z-value is above the

```{r, warning=FALSE, message = FALSE, echo = FALSE}
# my alpha for the z 
target_probability <- 1 - 0.05

#start here 
x <- 1

# two decimals should be fine
tolerance <- 1e-3

#find the z value using the pnorm function
while (abs(pnorm(x)) < target_probability ) {
  x <- x + tolerance  
}

# Adjust the step size based on your problem  
cat("critical value of z =", x, "corresponding to an alpha = 0.05." , "\n")

Second, we restricted the metaregression to studies using CBT and the FDA-approved antidepressants fluoxetine and escitalopram.

Third, we restricted the metaregression to studies that *did not* use waitlist as their control.

Fourth, we tested whether our results could be accounted for by differential regression to the mean. In an initial step, we tested in metaregression as above for differences in baseline means for those medication and psychotherapy studies that shared the same instruments. We then accounted for the effect that possible differing baseline values across all studies studies' outcomes by regressing the SMDs of studies on their respective baseline values of the instrument with which the SMD was calculated. For example, in a study that used the CDRS, this was used to calculate the SMD. This SMD was then used as the outcome and the baseline value of the CDRS, termed *BaselineScore* below, in that study was used as the independent variable as follows: $$
SMD = \beta_0 + \beta_1 BaselineScore + \epsilon
$$ {#eq-12}

This allows us to estimate the effect of the baseline score on the SMD by evaluating $\beta_1$. We can then re-arrange Equation 12 to extract the residuals, that is, the variability in the SMD that is not accounted for by the baseline score.

$$
\epsilon = SMD - (\beta_0 + \beta_1 BaselineScore) ) \
$$

where, $\epsilon$ is the adjusted SMD, for which we simulated standard errors and then used it the outcome in metaregression, as described above for the unadjusted SMDs.

@@

### Analysis of all studies

@@

z_psy_con_vs_med_con <- 0
for(i in 1: length(list_model_1_meta_reg)){
z_psy_con_vs_med_con[i] <- list_model_1_meta_reg[[i]]$zval[condition=="psychotherapy_control"]

@@

#### Addressing Regression to the Mean

We first conducted a meta-analysis to test for differences in means at baseline in the two instruments, CDRS-R and HAM-D on which there was a sufficient number of studies to metanalyse.

@@

### get CDRS studies
cdrs_studies_for_baseline_metareg <- list_cdrs_study[[1]] # created this above, you can take any of these, as the means and sds are repeated.

#turn sds to sec
cdrs_studies_for_baseline_metareg$baseline_stand_error <- cdrs_studies_for_baseline_metareg$baseline_sd/
                                        sqrt(cdrs_studies_for_baseline_metareg$baseline_n)

model_1 <- as.formula(~ four_level_var)
condition <-  levels(list_df_simulated[[1]]$four_level_var) # this is the four level var 
                                                              #throughout

results_cdrs_metareg_baseline <- run_metareg_and_extract(cdrs_studies_for_baseline_metareg,  
                                                         model_1, condition)

### get HAM-D studies
hamd_study_ids <-
  df_appl_v_orange[df_appl_v_orange$instrument_name == "hamd",]$new_study_id

list_hamd_study <- list_df_simulated[[1]][list_df_simulated[[1]]$new_study_id
                                                          %in% hamd_study_ids,]

hamd_studies_for_baseline_metareg <- list_hamd_study #

#turn sds to sec
hamd_studies_for_baseline_metareg$baseline_stand_error <- hamd_studies_for_baseline_metareg$baseline_sd/
                                        sqrt(hamd_studies_for_baseline_metareg$baseline_n)

model_1 <- as.formula(~ four_level_var)
condition <-  levels(list_df_simulated[[1]]$four_level_var) # this is the four level var 
                                                              #throughout
results_hamd_metareg_baseline <- run_metareg_and_extract(hamd_studies_for_baseline_metareg, 
                                                         model_1, condition)

results_hamd_metareg_baseline$condition <-  levels(list_df_simulated[[1]]$four_level_var)

@@

As can be seen in Table 3a and 3b, there baseline scores in medication studies were on average substantially higher for the CDRS (10.9 points for the control arms) and the HAM-D (7.3 points for the control arms), respectively. As can be seen in Supplementary Tables XX <!--# Charlotte, please create these out of results_cdrs_metareg_baseline[[1]]  and the same for HAM-D--> the differences were significant at p\<0.05 for CDRS, thought they did not reach statistical significance for the HAM-D (p = 0.0573). <!--# Charlotte, please add the n for each row. Easiest to modify the function so that it appends the baseline n together with the condition. -->

```{r, warning=FALSE, message = FALSE, echo=FALSE}
knitr:: kable(results_cdrs_metareg_baseline[[2]], caption="Table 3a CDRS baselilne scores")
```

```{r, warning=FALSE, message = FALSE, echo=FALSE}
knitr:: kable(results_hamd_metareg_baseline[[2]], caption="Table 3 HAM-D baselilne scores")
```

We next addressed potential regression to the mean by residualising SMDs by the baseline scores in one regression model.

```{r, warning=FALSE, message = FALSE, echo=FALSE}
#### Let's start from the beginning and regress cohens on the baseline dataframe.

# create a new list with dfs for the sims
df_long_for_metan_adj_cohens_d <- df_long_for_metan
# rename the cohens_d value to make space for the new cohen's d
df_long_for_metan_adj_cohens_d <-
  df_long_for_metan_adj_cohens_d %>% 
rename(unadjusted_cohens_d = cohens_d,
) 
# estimate the cohen's d from the residuals
model_adjust_cohens_d <-  lm(df_long_for_metan_adj_cohens_d$unadjusted_cohens_d ~ df_long_for_metan_adj_cohens_d$baseline_mean, na.action=na.exclude)

# there is a great effect of the baseline means, as shown here
results_adjusted_cohens_d <- broom:: tidy(model_adjust_cohens_d )

# extract the residuals which are the adjusted cohen's d, to get the simulated standard errors
df_long_for_metan_adj_cohens_d$cohens_d <- residuals(model_adjust_cohens_d )

# apply function to simulate standard errors
list_df_simulated_adj_cohens_d <- simulate_dataframes_for_st_errors(df = df_long_for_metan_adj_cohens_d, 
                                   num_repetitions = 1000, 
                                   seed  = 1974,
                                   n = length(unique(df_long_for_metan_adj_cohens_d$new_study_id)), 
                                   a = 0.45, 
                                   b = .9, 
                                   mean = 0.65, 
                                   sd = 0.2)

# add the four level variable and re-level
for(i in 1: length(list_df_simulated_adj_cohens_d)){

  list_df_simulated_adj_cohens_d[[i]]$arm_effect_size <- factor(list_df_simulated_adj_cohens_d[[i]]$arm_effect_size)
  list_df_simulated_adj_cohens_d[[i]] <- list_df_simulated_adj_cohens_d[[i]] %>% 
    mutate(four_level_var = case_when(psy_or_med == 0 & arm_effect_size == "cohens_d_active" ~ "medication_active",
                                      psy_or_med == 0 & arm_effect_size == "cohens_d_control" ~ "medication_control",
                                      psy_or_med == 1 & arm_effect_size == "cohens_d_active" ~ "psychotherapy_active",
                                      psy_or_med == 1 & arm_effect_size == "cohens_d_control" ~ "psychotherapy_control")
    )

  list_df_simulated_adj_cohens_d[[i]]$four_level_var <- factor(list_df_simulated_adj_cohens_d[[i]]$four_level_var) # turn to factor

  # relevel so that medication control becomes the reference category for the regression
  #  list_df_simulated[[i]]$four_level_var <- relevel(list_df_simulated[[1]]$four_level_var, ref = "medication_control")
  list_df_simulated_adj_cohens_d[[i]]$four_level_var <-   fct_relevel(list_df_simulated_adj_cohens_d[[i]]$four_level_var,
                                                         "medication_control",
                                                         "medication_active", 
                                                         "psychotherapy_control",
                                                         "psychotherapy_active")
}

# Run metaregression model (assuming that the four level variable is present given that I create this dataframe from the list_)
# specify model
model_1 <- as.formula(~ four_level_var)

# run metareg function 
# here you can use the simulated results directly
list_model_1_meta_reg_adj_cohens_d <- run_metaregression(list_df_simulated_adj_cohens_d, model_1)

# extract coefficients and model characteristics.
condition <- levels(list_df_simulated_adj_cohens_d[[1]]$four_level_var)
aggregate_results_overall_adjusted_cohens_d <-  aggregate_model_results(list_model_1_meta_reg_adj_cohens_d, condition)
aggregate_results_overall_adjusted_cohens_d <- cbind(aggregate_results_overall_adjusted_cohens_d, condition)

# extract SMDs from the  list
list_dummy_var_means_adjusted_cohens_d <-  lapply(list_model_1_meta_reg_adj_cohens_d, extract_coefficients_func)

# calculate mean SMDs and ses
df_mean_coefs_from_sim_adjusted_cohens_d <- calculate_mean_coefs_ses(list_dummy_var_means_adjusted_cohens_d, condition)

df_mean_coefs_from_sim_adjusted_cohens_d <- data.frame(cbind(condition, df_mean_coefs_from_sim_adjusted_cohens_d))

As can be seen in Table 4, the baseline mean has a very strong association with the SMD.

```{r, warning=FALSE, message = FALSE, echo=FALSE}
knitr:: kable(results_adjusted_cohens_d

              , caption = "Table 4. Effect of baseline values on SMDs")
```

Table 5 shows the adjusted SMDs with their confidence intervals that we estimated in metaregression. As can be inferred from the non-overlapping confidence intervals, the difference between the two control conditions was significant (details of the regression model can be found in Supplementary Table XX <!--# Charlotte, please add there the aggregate_results_overall_adjusted_cohens_d -->). Differences were significant between active medication and psychotherapy control, but not between any of the other groups.

```{r, warning=FALSE, message = FALSE, echo=FALSE}
knitr:: kable (df_mean_coefs_from_sim_adjusted_cohens_d, caption = "Table 5. SMDs Adjusted by baseline score" )
```

Finally, we tested whether the simulated correlation values we used for simulation of the standard errors had an appreciable effect on the values of the adjusted SMDs. Supplemental Figure X shows that the variability of the estimated z-value was minimal<!--# Charlotte, the figure below should be supplemental-->.

# plot the mean z for the difference between controls
z_psy_con_vs_med_con <- 0
for(i in 1: length(list_model_1_meta_reg_adj_cohens_d )){
  z_psy_con_vs_med_con[i] <- list_model_1_meta_reg_adj_cohens_d [[i]]$zval[condition=="psychotherapy_control"]

stability_sims_adjusted_cohens_d <- data.frame(n_sims = 1:1000, z_value = z_psy_con_vs_med_con)

stab_sims_adjusted_cohens_d <- stability_sims_adjusted_cohens_d %>%
  ggplot(aes(x = n_sims, y = z_value)) +
  geom_point() +
  geom_hline(yintercept = 1.65, linetype = "dotted", color = "red") +
  geom_segment(aes(x = 300, xend = 300, y = 2, yend = 1.65),
               arrow = arrow(length = unit(0.3, "cm")),
               color = "black") +
  annotate("text", x = 300, y = 2.0, label = "p < 0.05 threshold (values above line significant)", hjust = -0.1, vjust = 0) +
  annotate("text", x = 300, y = stability_sims_adjusted_cohens_d$z_value[1] - 0.1, label = "simulated z-values",
           hjust = -0.1, vjust = 0) +
  theme_minimal() +
  labs(
    x = "Number of Simulations",
    y = "z-value",
    title = "Stability of the Statistic of the Difference between \nMedication and Psychotherapy Control Arms",
    subtitle = "Results from 1000 simulations with within-group standard errors"
  )
stab_sims_adjusted_cohens_d
pdf(file = "stab_sims_adjusted_cohens_d.pdf")
stab_sims_adjusted_cohens_d
dev.off()
